Distributional Trees and Forests for Circular Data with an Application to Probabilistic Wind Direction Forecasting

### 2019-10-20 pkg/circtree/inst/wind_paper/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2803 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5526f6db8a1ab44ea03cd4f8f2a21594e5e3b2a8

\Abstract{
As circular data can be found in many different subject areas accounting for periodicity has 
become an important aspect in establishing appropriate statistical models.
For probabilistic modeling of circular data the von Mises distribution is widely used.
To capture how its parameters change with covariates, a regression tree model
is proposed as an alternative to more commonly-used additive models. The resulting
distributional trees are easy to interpret, can detect non-additive effects, and select
covariates and their interactions automatically. In order to stabelize and regularize
the model an ensemble of distributional trees is combined yielding a distributional
forest which allows for modeling smooth effects as well.
For illustration, hourly wind direction forecasts are obtained at two airports in Austria
(Vienna and Innsbruck) based on a set of meteorological measurements from each of the 
airports and surrounding weather stations.
}
\Keywords{Distributional Trees; Circular Response; Von Mises Distribution}

@@

In many approaches to model circular data it is assumed that the circular variable of interest 
follows a circular distribution, in particular the von Mises distribution which is also known 
as the ``circular normal distribution'' (see Section~\ref{sec:prob_circ} for more details). 
This was also the basis which early regression models for circular data were built on as they 
regress the distribution parameter(s) on covariates.

One of the first regression models with a circular response variable and linear covariates
was presented by \cite{Gould:1969} where the circular mean, hence the location parameter of 
the von Mises distribtuion, is predicted by a linear combination of covariates.
\cite{Johnson+Wehrly:1978} took this idea one step further by employing generalized linear 
model (GLM)-type approach, i.e., by plugging in a link function such that the values of the 
linear predictor are transformed to a restricted interval of length $2\pi$. 
This model was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
introducing another specific transformation function which maps the values of the real line to the 
unit circle. Moreover, the presented methodology allows for both distribution parameters, the mean direction and the concentration parameter, to be modeled not only in separate models but also in a 
mixed model. There, within an iterative optimization algorithm the two parameters are estimated and 
improved stepwise until convergence. To obtain proper initial values it is recommended to apply
two separate models where either one of the parameters is fixed while the other is modeled depending
on covariates. While providing a well elaborated theory this methodology and the contained optimization 
problem have shown to be very challenging, mainly due to the complexity encountered in optimizing the 
log-likelihood function that is not globally concave, and thus bear a high risk of not converging if 
not applied thoughtfully \citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}.
An alternative approach to regression modeling of circular data that avoids this strong dependence 
on choosing well-fitting initial values is presented by \cite{Mulder+Klugkist:2017} employing a GLM 
in a Bayesian framework.

The named methods and also many of their further developments 
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive regression models.
While they provide a powerful toolbox, particularly for modeling smooth effects, they may not be 
able to capture all types of effects. For example, abrupt changes might not be detected properly owing 
to the fixed structure enforced by model assumptions such as the type of link function and the 
form of linear combinations.

Therefore, we propose a very flexible alternative approach by adapting regression trees and random 
forests to circular data, specifically by employing distributional trees and distributional forests
following \cite{Schlosser+Hothorn+Stauffer:2019}. 
In that way no model assumptions are required as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while combining an 
ensemble of trees to a forest model allows for modeling smooth changes as well. 
In that way, a high level of flexibility is provided while a full distributional model is specified 
by distributional trees and forests such that a wide range of inference is made available.
These features and the corresponding methodology are presented in Section~\ref{sec:tree_forest} after
discussing distributional models for circular data in Section~\ref{sec:prob_circ}.
In Section~\ref{sec:wind} the presented models a applied to probabilistic forecasting of wind direction
at the airports of Innsbruck and Vienna.

@@

Employing the score function
\begin{equation}
\begin{aligned}
s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
\smallskip &,
\quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y_i) \bigg) 
\\
&= \bigg(\kappa \sin(y-\mu)
\smallskip &,
\quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
\end{aligned}
\end{equation}
the optimization problem in Equation~\ref{eq:optim} can also be defined by 
%setting the sum of the first derivative of the log-likelihood function evaluated for all observations to zero, i.e., by setting
\begin{equation}
\sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
\end{equation}

\begin{figure}[t]
\centering
\includegraphics[width=4cm]{density_vM.pdf}
\caption{Empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand).
}
\label{fig:density_vM} 
\end{figure}

Figure~\ref{fig:density_vM} illustrates a distributional model fitted to circular data
employing the von Mises distribution.
% on observed wind direction data from the airport of Innsbruck.
This example shows that reasonably well fitting probabilistic models can
be obtained for circular data by specifying an appropriate distribution family 
such as the von Mises distribution and estimating its parameters via maximum likelihood. 
However, this procedure considers only the response variable $Y$. Including covariates 
can improve any model remarkably as they might provide valuable additional information.
Therefore, distributional regression models are of high interest for all types of data,
including circular data.

\section{Regression Trees and Forests for Circular Data}
\label{sec:tree_forest}
%\subsection{Motivation}
A common way to including covariates as regressors in distributional models is to apply
generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} where each distribution parameter is estimated
by its own generalized additive model (GAM).
As already discussed in Section~\ref{sec:introduction}, additive strategies work well 
for smooth effects but difficulties can arise in case of non-additive changes. Additionally, 
specifying a model properly can be very challenging, particularly for a high number of covariates 
and no information about possible interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

\subsection{Distributional Trees for Circular Data}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly for 
complex data with strong variations. Therefore, separating the data set into more homogeneous 
subgroups based on covariates before fitting a local model in each of these subgroups allows 
to capture (possibly) group specific effects more precisely and hence can result in a better 
fitting overall model.
This is the general idea of regression trees which are combined with distributional regression
modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the unbiased recursive partitioning 
algorithms MOB \citep{Zeileis+Hothorn+Hornik:2008} or CTree \citep{Hothorn+Hornik+Zeileis:2006}.
In the presented distributional trees a full distributional model is specified in each node of the 
tree. For this purpose a distribution family has to be chosen in advance. Hence, selecting the 
von Mises distribution allows for an application of distributional trees to circular data following 
the tree building algorithm as introduced by \cite{Schlosser+Hothorn+Stauffer:2019}.

In particular, in each node the crucial decision on how and where to split the data is based on model scores which are obtained by evaluating the score function $s$ at the individual observations and 
parameter estimates. For the von Mises distribution and a data set of $n$ observations
this yields an $n \times 2$ matrix that can be employed as a kind of residual, capturing how well
each given observation conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, 
respectively.
To capture dependence on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test (MOB) or a permutation test (CTree).
In each partitioning step, the covariate with the highest significant association (i.e., lowest
significant $p$-value, if any) is selected for splitting the data. The corresponding split point
is chosen either by optimizing the log-likelihood (MOB) or a two-sample test statistic (CTree)
over all possible partitions.
%and the covariate corresponding to the strongest change is selected as split varialbe.
%Next a split point is chosen within the selected covariate by maximizing a partitioned likelihood.
This procedure is repeated recursively until there are no significant parameter instabilities or until another stopping criterion is met (e.g., subgroup size or tree depth).

@@

While tree models can capture non-additive effects they do not allow for purely smooth effects and therefore might impose abrupt changes even in case of rather smooth transitions. This can be avoided
by combining an ensemble of trees in order to obtain a more stable forest model.

\subsection{Distributional Forests for Circular Data}
\label{sec:circforest}
A natural extension of distributional trees for circular data are ensembles or forests 
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by \cite{Breiman:2001} average the predictions of an ensemble
of trees, each built on a subsample or bootstrap of the original data. 
A generalization of this strategy is to obtain weighted predictions
by adaptive local likelihood estimation of the distributional parameters
\citep{Schlosser+Hothorn+Stauffer:2019, Hothorn+Zeileis:2017}. 
More specifically, for each possibly new observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights $w_i^{forest}(\bm{z})$ 
%\citep{Lin+Jeon:2006}
is obtained that is based on the number of trees in which $\bm{z}$ is assigned to the same terminal 
node as each learning observation $y_i, i \in \{1,\ldots,n\}$.
Hence, for a forest of $T$ trees the weights are calculated via
\begin{equation}
w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
\frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree.
Hence, similar observations ending up more often in the same terminal node have higher weights
and in that way a stronger influence in the estimation process.

@@

\subsection{Data}
For both airports, Innsbruck and Vienna, observations of various meteorological quantities 
such as wind direction, temperature, air pressure and humidity
measured either directly at the airports or at surrounding weather stations are used as 
covariates in order to predict the wind direction at the corresponding airport 
in the following hour.
Next to the current measurements at each of the weather stations also differences in these 
quantities between the considered stations are included. For example, differences in air 
pressure between Innsbruck and either a western or eastern location are expected to have a 
high influence on wind directions. Moreover, changes of these quantities over time are 
included as well as daytime. In total, 550 covariates are provided for Innsbruck and 522 
covariates for Vienna. 
A list of all covariates together with their abbreviations as used in this paper
can be found in Table~\ref{tab:covariates}.
The data sets each provide observations from 5 years, in particular from 2014 to 2017, both included.
After eliminating observations with at least one missing value a total number of 26861 time points
for Innsbruck and 33853 for Vienna are available.
As winds with very low speed tend to vary their direction strongly they are not expected to be very 
useful/meaningful for predictions. Moreover, the performance and the stability of the employed reference 
models might suffer from the influence of these ``noise'' observations. Therefore, only observations
corresponding to a wind speed of at least $1$ m/s are considered.

@@

\subsection{Evaluation Setting and Models Applied}
Along with distributional trees and forests for circular data three other models have been applied
in this study. In particular, they have been evaluated in a five-fold cross validation with each of 
the five years contained in the data set being used as testing data once after the models where 
learned on the remaining four years. 
In order to compare the performances of the employed models a circular version of the continuous 
ranked probability score (CRPS) as introduced by \cite{Grimit+Gneiting+Berrocal:2006} is calculated.
Just as the linear version of the CRPS \citep[see][for more details]{Hersbach:2000} it measures the 
difference between an observation and the corresponding predicted distribution function in order 
to assess the probabilistic goodness of fit of the assessed model. Hence, the lower the CRPS value 
the better the performance. In order to adjust the general methodology of the CRPS to circular data 
\cite{Grimit+Gneiting+Berrocal:2006} employ an auxiliary function which measures the angular distance 
between two circular variables. 
The circular CRPS is evaluated out of sample and the resulting values are averaged over 
each observation contained in the testing data and the corresponding predicted distribution 
parameters of each of the models.
To measure the improvement of the trees, forests, persistence and linear models 
over the climatology model the skill score based on the circular CRPS is evaluated
with the climatology model used as reference:
\begin{equation}
\text{CRPSS}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{climatology}}}.
\end{equation}

To test whether distributional trees and forests can be applied as a reasonable alternative to already
existing approaches to modeling circular data, particularly in the field of meteorological forecasting,
three commonly used models are selected as reference models.
The exact settings of the tree and forest models as well as a specification of the reference models
is provided in the remainder of this section.

@@

\paragraph{Distributional Forest}
Following the description in Section~\ref{sec:circforest} a distributional forest is constructed.
As for the distributional tree all available covariates are used to learn a distributional 
forest of 100 trees. Each of these trees is built by the CTree algorithm on a subsample 
containing a fraction of 30 percent of the original learning data. As it can be expected that 
the lagged response variable (as included as a covariate) has a strong influence on the 
response variable it is desirable to employ it in each tree of the forest. Therefore, all 
covariates are considered for each tree
(by setting the argument \code{mtry} of our implementation to the total number of covariates)
in order to guarantee for the lagged response to be available.
Since a high number of possible split points lead to a high computational effort that is 
particularly noteworthy for forest models the covariates are binned allowing for a maximum of
$50$ classes.
% control = disttree_control(nmax = c("yx" = Inf, "z" = 50)))
As stopping criteria the minimal number of observations to perform a split (\texttt{minsplit}) 
is set to 20, the minimal number of observations in each terminal node (\texttt{minbucket})
is set to 7, and the significance level for variable selection (\texttt{alpha}) is kept at its 
maximum value of 1 in order to build trees as large as possible as commonly done for forest models.

@@

To establish a distributional model of persistence in this study a set of six observations
is considered to make a prediction for a selected day and time point. In particular,
the six previous time points are considered. In order to allow for a stronger influence of 
observations closer to the prediction time point exponential smoothing is employed, i.e.,
exponentially decaying weights are included with a smoothing factor of $0.5$ such that
in each prediction step the current prediction and the previous observation both have an 
equal influence rate of 50 percent. If a seventh or even earlier observations were considered 
as well their weights would be $<0.01$ and thus are not included anymore to simplify calculations.
For this selected set of weighted observations the distribution parameters of the von Mises 
distribution can be estimated as presented in Section~\ref{sec:prob_circ} such that a full
parametric model is specified allowing for probabilistic predictions and an evaluation of the CRPS.
As this approach only considers data from the previous six days a cross validation is not 
applicable, however, the same evaluation framework as for the other four models can be employed 
by obtaining predictions only for the testing data of the cross validation without using 
the corresponding learning data.

\paragraph{Climatology}
If there are no climatological changes within a certain number of years it can be expected that 
for a selected day of the year weather conditions will always be the same over all years 
considered.
Again, this idea provides a very simple and stable way of obtaining predictions for a considered 
meteorological variable of interest that can be very useful but its performance can also suffer 
dramatically from a high variance and abrupt changes, especially if they occur frequently during one day.

@@

Similar to the applied persistence model observations are selected to fit a distributional climatology
model. For a specified day and time point observations from the exact same time point and the same day 
as well as the 30 surrounding days from all other years contained in the learning data are selected and
parameters are estimated following again Section~\ref{sec:prob_circ}.

\paragraph{Generalized Linear Model}
Another commonly applied approach is to linearly regress the circular variable of interest on
its lagged values. For the here discussed application this corresponds to establishing a linear model
with wind direction as the outcome variable and wind direction of the previous time point as covariate.
Following \cite{Mulder+Klugkist:2017}, the employed model plugs in a link function in order
to guarantee values on the interval $[-\pi,\pi]$ and is of the form
\begin{align}
dd = \beta_0 + g(\beta_1 \cdot dd_{lagged})
\end{align}
with $\beta_0$ being a circular intercept, $\beta_1$ the regression coefficient and 
$g$ the link function $g(x) = r \cdot \arctan(x)$ with an optional factor $r$. 
Using the corresponding implementation provided in the \textsf{R} package \textbf{circglmbayes}
by \cite{Mulder+Klugkist:2017} the function \textbf{circGLM} estimates the
intercept and regression coefficient along with the the concentration parameter $\kappa$.

\fixme{add references and further information on the estimation process within circGLM}

@@

@@

>>>>

\Abstract{
As circular data can be found in many different subject areas accounting for periodicity has 
become an important aspect in establishing appropriate statistical models.
For probabilistic modeling of circular data the von Mises distribution is widely used.
To capture how its parameters change with covariates, a regression tree model
is proposed as an alternative to more commonly-used additive models. The resulting
distributional trees are easy to interpret, can detect non-additive effects, and select
covariates and their interactions automatically. In order to stabelize and regularize
the model an ensemble of distributional trees is combined yielding a distributional
forest which allows for modeling smooth effects as well.
For illustration, hourly wind direction forecasts are obtained at two airports in Austria
(Vienna and Innsbruck) based on a set of meteorological measurements from each of the 
airports and surrounding weather stations.
}
\Keywords{Distributional Trees and Forests; Circular Response; Von Mises Distribution}

@@

In many approaches to modeling circular data it is assumed that the circular variable of interest 
follows a circular distribution, in particular the von Mises distribution which is also known 
as the ``circular normal distribution'' (see Section~\ref{sec:prob_circ} for more details). 
This is also the basis which early regression models for circular data are built on as they 
regress the parameter(s) of the von Mises distribution on covariates.

One of the first regression models with a circular response variable and linear covariates
was presented by \cite{Gould:1969} where the circular mean, hence the location parameter of 
the von Mises distribtuion, is predicted by a linear combination of covariates.
\cite{Johnson+Wehrly:1978} took this idea one step further by employing a generalized linear 
model (GLM)-type approach, i.e., by plugging in a link function such that the values of the 
linear predictor are transformed to a restricted interval of length $2\pi$. 
This model was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
introducing another specific transformation function which maps the values of the real line to the 
unit circle. Moreover, the presented methodology allows for both distribution parameters, the mean direction and the concentration parameter, to be modeled not only in separate models but also in a 
mixed model. There, within an iterative optimization algorithm the two parameters are estimated and 
improved stepwise until convergence. To obtain proper initial values it is recommended to apply
two separate models where either one of the parameters is fixed while the other is modeled depending
on covariates. While being built on well elaborated theory this methodology and the contained 
optimization problem have shown to be very challenging in their application, mainly due to the 
complexity encountered in optimizing the log-likelihood function that is not globally concave, 
and thus bear a high risk of not converging if not applied thoughtfully 
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}.
An alternative approach to regression modeling of circular data that avoids this strong dependence 
on choosing well-fitting initial values is presented by \cite{Mulder+Klugkist:2017} employing a GLM 
in a Bayesian framework.

The named methods and also many of their further developments 
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive regression models.
While they provide a powerful toolbox for modeling circular data, particularly in case of
smooth effects, they may not be able to capture all types of effects. For example, abrupt changes 
might not be detected properly owing to the enforced additive structure and model assumptions 
such as the type of link function and the form of linear combinations.

Therefore, we propose a very flexible alternative approach by adapting regression trees and random 
forests to circular data, specifically by employing the von Mises distribution within the methodology
of distributional trees and distributional forests following \cite{Schlosser+Hothorn+Stauffer:2019}. 
In that way no model assumptions are required as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while combining an 
ensemble of trees to a forest model allows for modeling smooth changes as well. 
Thus, a high level of flexibility is provided while a full distributional model is specified 
by distributional trees and forests such that a wide range of inference is made available.
The corresponding methodology and its features are presented in Section~\ref{sec:tree_forest} after
discussing distributional models for circular data in Section~\ref{sec:prob_circ}.
In Section~\ref{sec:wind} the presented models are applied to probabilistic forecasting of wind direction
at the airports of Innsbruck and Vienna.

@@

By means of the score function
\begin{equation}
\label{eq:scores}
\begin{aligned}
s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
\smallskip &,
\quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y_i) \bigg) 
\\
&= \bigg(\kappa \sin(y-\mu)
\smallskip &,
\quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
\end{aligned}
\end{equation}
an indicator of deviation and in that way a measure of goodness of fit of the model can be obtained 
for each observation $y_i$. Moreover, the optimization problem in Equation~\ref{eq:optim} can also 
be defined by 
%setting the sum of the first derivative of the log-likelihood function evaluated for all observations to zero, i.e., by setting
\begin{equation}
\sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
\end{equation}

\begin{figure}[t]
\centering
\includegraphics[width=4cm]{density_vM.pdf}
\caption{Illustration of a distributional model fitted to circular data by maximum likelihood-estimation
employing the von Mises distribution. Empirical histogram (gray) and fitted density (red line) are depicted
along with the estimated location parameter (red hand).
}
\label{fig:density_vM} 
\end{figure}

Figure~\ref{fig:density_vM} illustrates a distributional model fitted to circular data
employing the von Mises distribution.
% on observed wind direction data from the airport of Innsbruck.
This example shows that reasonably well fitting probabilistic models can
be established for circular data by specifying an appropriate distribution family 
such as the von Mises distribution and estimating its parameters via maximum likelihood. 
However, this procedure considers only the response variable $Y$. Including covariates 
can improve any model remarkably as they might provide valuable additional information.
Therefore, distributional regression models are of high interest for all types of data,
including circular data.

\section{Regression Trees and Forests for Circular Data}
\label{sec:tree_forest}
%\subsection{Motivation}
A common way of including covariates as regressors in distributional models is to apply
generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} where each distribution parameter is estimated
by its own generalized additive model (GAM).
As already discussed in Section~\ref{sec:introduction}, additive strategies work well 
for smooth effects but difficulties can arise in case of non-additive changes. Additionally, 
specifying a model properly can be very challenging, particularly for a high number of covariates 
and no information on possible interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

\subsection{Distributional Trees for Circular Data}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly for 
complex data with strong variations. Therefore, separating the data set into more homogeneous 
subgroups based on covariates before fitting a local model in each of these subgroups allows 
to capture (possibly) group specific effects more precisely and hence can result in a better 
fitting overall model.
This is the general idea of regression trees which are combined with distributional regression
modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the unbiased recursive partitioning 
algorithms MOB \citep{Zeileis+Hothorn+Hornik:2008} or CTree \citep{Hothorn+Hornik+Zeileis:2006}.
By applying the presented distributional trees a full distributional model is specified in each 
node of the tree. For this purpose a distribution family has to be chosen in advance. Hence, 
selecting the von Mises distribution allows for an application of distributional trees to circular
data following the tree building algorithm as introduced in Section~2.2. of 
\cite{Schlosser+Hothorn+Stauffer:2019}.

In particular, in each node the crucial decision on how and where to split the data is based on model
scores which are obtained by evaluating the score function $s$ as defined in Equation~\ref{eq:scores} 
at the individual observations and parameter estimates. For the von Mises distribution and a data set 
of $n$ observations this yields an $n \times 2$ matrix that can be employed as a kind of residual,
capturing how well each given observation conforms with the estimated location $\hat{\mu}$ and precision
$\hat{\kappa}$, respectively.
To capture dependence on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test~(MOB) or a permutation test~(CTree).
In each partitioning step, the covariate with the highest significant association (i.e., lowest
significant $p$-value, if any) is selected for splitting the data. The corresponding split point
is chosen either by optimizing the log-likelihood~(MOB) or a two-sample test statistic~(CTree)
over all possible partitions.
%and the covariate corresponding to the strongest change is selected as split varialbe.
%Next a split point is chosen within the selected covariate by maximizing a partitioned likelihood.
This procedure is repeated recursively until there are no significant parameter instabilities or until another stopping criterion is met (e.g., subgroup size or tree depth).

@@

While tree models can capture non-additive effects their structure and the consequential strict 
separation of data into subgroups hinders an adequate depiction of smooth effects. Therefore, 
they might impose abrupt changes even in case of rather smooth transitions. This can be avoided 
by combining an ensemble of trees in order to obtain a more stable forest model.

\subsection{Distributional Forests for Circular Data}
\label{sec:circforest}
A natural extension of distributional trees for circular data are ensembles or forests 
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by \cite{Breiman:2001} average the predictions of an ensemble
of trees, each built on a subsample or bootstrap of the original data. 
A generalization of this strategy is to obtain weighted predictions by adaptive local 
likelihood estimation of the distributional parameters 
\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019, Hothorn+Zeileis:2017}. 
More specifically, for each possibly new observation~$\bm{z}$ a set of averaged ``nearest neighbor''
weights $w_i^{forest}(\bm{z})$ 
%\citep{Lin+Jeon:2006}
is obtained that is based on the number of trees in which $\bm{z}$ is assigned to the same terminal 
node as each learning observation $y_i, i \in \{1,\ldots,n\}$.
Hence, for a forest of $T$ trees the weights are calculated via
\begin{equation}
w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
\frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree.
Hence, similar observations ending up more often in the same terminal node have higher weights
and in that way a stronger influence in the estimation process.

@@

\subsection{Data}
For both airports, Innsbruck and Vienna, observations of various meteorological quantities 
such as wind direction, temperature, air pressure and humidity
measured either directly at the airports or at surrounding weather stations are used as 
covariates in order to predict the wind direction at the corresponding airport 
in the following hour.
Next to the current measurements at each of the weather stations also differences in these 
quantities between the considered stations are included. For example, differences in air 
pressure between Innsbruck and either a western or eastern location are expected to have a 
high influence on wind directions. Moreover, changes of these quantities over time are 
included as well as daytime. In total, 550 covariates are provided for Innsbruck and 522 
covariates for Vienna. 
A list of all covariates together with their abbreviations as used in this paper
can be found in Table~\ref{tab:covariates}.
The data sets each provide observations from 5 years, in particular from 2014 to 2017, both included.
After eliminating observations with at least one missing value a total number of 26861 time points
for Innsbruck and 33853 for Vienna are available.
As winds with very low speed tend to vary their direction rapidly and almost randomly they are not 
expected to have a meaningful impact on predictions. Moreover, the performance and the stability 
especially of the employed reference models might suffer from the influence of these ``noise'' 
observations. Therefore, only observations corresponding to a wind speed of at least $1$ m/s are 
considered.

@@

\subsection{Evaluation Setting and Models Applied}
Along with distributional trees and forests for circular data three other models have been applied
in this study. In particular, they have been evaluated in a five-fold cross validation with each of 
the five years contained in the data set being used as testing data once after the models where 
learned on the remaining four years.

In order to compare the performances of the employed models a circular version of the continuous 
ranked probability score (CRPS) as introduced by \cite{Grimit+Gneiting+Berrocal:2006} is calculated.
Just as the linear version of the CRPS \citep[for more details see][]{Hersbach:2000} it measures the 
difference between an observation and the corresponding predicted distribution function in order 
to assess the probabilistic goodness of fit of the fitted model. Hence, the lower the CRPS value 
the better the performance. In order to adjust the general methodology of the CRPS to circular data 
\cite{Grimit+Gneiting+Berrocal:2006} employ an auxiliary function which measures the angular distance 
between two circular variables.

The circular CRPS is evaluated out of sample and the resulting values are averaged over 
each observation contained in the testing data and the corresponding predicted distribution 
parameters of each of the models.

\fixme{Comment on aggregation}

To measure the improvement of one of the applied methods over a climatology model (which will 
be described in detail subsequently) the skill score based on the circular 
CRPS of the assessed method is evaluated with the climatology model used as reference:
\begin{equation}
\text{CRPSS}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{climatology}}}.
\end{equation}

Three commonly used methods are selected as reference models in order to investigate whether 
distributional trees and forests can be applied as a reasonable alternative to already
existing approaches to modeling circular data, particularly in the field of meteorological forecasting.
The exact settings of the tree and forest models as well as a specification of the reference models
is provided in the remainder of this section.

@@

\paragraph{Distributional Forest}
Following the description in Section~\ref{sec:circforest} a distributional forest is constructed.
As for the distributional tree all available covariates are used to learn a distributional 
forest of 100 trees. Each of these trees is built by the CTree algorithm on a subsample 
containing a fraction of 30 percent of the original learning data. As it can be expected that 
the lagged response variable (as included as a covariate) has a strong influence on the 
response variable it is desirable to employ it in each tree of the forest. Therefore, all 
covariates are considered for each tree
(by setting the argument \code{mtry} of our implementation to the total number of covariates)
in order to guarantee for the lagged response to be available.
Since a high number of possible split points leads to a high computational effort that is 
particularly noteworthy for forest models the covariates are binned allowing for a maximum of
$50$ classes. In that way the number of split points that have to be considered is reduced
in a reasonably.
% control = disttree_control(nmax = c("yx" = Inf, "z" = 50)))
As stopping criteria the minimal number of observations to perform a split (\texttt{minsplit}) 
is set to 20, the minimal number of observations in each terminal node (\texttt{minbucket})
is set to 7, and the significance level for variable selection (\texttt{alpha}) is kept at its 
maximum value of 1 in order to build trees as large as possible as commonly done for forest models.

@@

To establish a distributional model of persistence in this study a set of six observations
is considered to make a prediction for a selected day and time point. In particular,
the six previous time points are considered. In order to allow for a stronger influence of 
observations closer to the prediction time point exponential smoothing is employed, i.e.,
exponentially decaying weights are included with a smoothing factor of $0.5$ such that
in each prediction step the current prediction and the previous observation both have an 
equal influence rate of 50 percent. If a seventh or even earlier observations were considered 
as well their weights would be $<0.01$ and thus are not included anymore to simplify calculations.
For this selected set of weighted observations the distribution parameters of the von Mises 
distribution can be estimated as presented in Section~\ref{sec:prob_circ} such that a full
parametric model is specified allowing for probabilistic predictions and an evaluation of the 
circular CRPS.
As this approach only considers data from the previous six days a cross validation is not 
applicable, however, the same evaluation framework as for the other four models can be employed 
by obtaining predictions only for the testing data of the cross validation without using 
the corresponding learning data.

\paragraph{Climatology}
If there are no climatological changes within a certain number of years it can be expected that 
for a selected time point and day of the year weather conditions will always be the same over all 
years considered.
Again, this idea provides a very simple and stable way of obtaining predictions for a considered 
meteorological variable of interest that can be very useful but its performance can also suffer 
dramatically from a high variance and abrupt changes, especially if they occur frequently during one day.

@@

Similar to the applied persistence model a set of observations is selected to fit a distributional
climatology model. For a specified day and time point observations from the exact same time point of 
the same day as well as of the 30 surrounding days from all other years contained in the learning data 
are selected and parameters are estimated following again Section~\ref{sec:prob_circ}.

\paragraph{Generalized Linear Model}
Another commonly applied approach is to linearly regress the circular variable of interest on
its lagged values. For the here discussed application this corresponds to establishing a generalized
linear model with wind direction as the outcome variable and wind direction of the previous time point 
as covariate. 
Following \cite{Mulder+Klugkist:2017} the employed model plugs in a link function in order
to guarantee values on the interval $[-\pi,\pi]$ and is of the form
\begin{align}
dd = \beta_0 + g(\beta_1 \cdot dd_{lagged})
\end{align}
with $\beta_0$ being a circular intercept, $\beta_1$ the regression coefficient and 
$g$ the link function $g(x) = r \cdot \arctan(x)$ with an optional factor $r$. 
The intercept and the regression coefficient are estimated in a Bayesian framework where an 
MCMC algorithm is used to obtain posterior samples. Moreover, a constant concentration
parameter is fitted to the full learning sample.

\fixme{add references and further information on the estimation process within circGLM?}

@@

Lisa Schlosser recieved a PhD scholarship granted from the University of Innsbruck.

@@

For the circular GLM considered as reference model the corresponding implementation is provided 
in the \textsf{R} package \textbf{circglmbayes} by \cite{Mulder+Klugkist:2017}. In particular the 
function \textbf{circGLM} is applied to estimates the intercept and regression coefficient along with 
the concentration parameter $\kappa$.
